The Current State of R Tools for Nonlinear Least Squares Modeling

### 2023-07-11 RefactoringNLS/revision202307/ReviewResponses.txt

Arkajyoti responses -- delete before submission:

I have read through your comments on the suggestions of the two referees and I agree with them. As usual, the document is well-organized, methodical, and nice way to check all the boxes.

For “Response 2:10“, I don’t think any link would be a good choice based on the referees’ concern. An alternative article (maybe the reference manual or vignette of the corresponding package  in question, like for unit testing) may serve as a viable replacement.
    Agree that no link is likely to be satisfactory. Modified to suggest search to find
    articles.

Some comments:

    Maybe cite your nlsr package, as “our” nlsr seems confusing? -- DONE

    There is some overlap between the Abstract and scope of our comparison, and I suggest making the 
        latter more succinct (for example the paragraph before TOOLS CONSIDERED)?
        DONE -- simplified.

    Maybe talking about referees’ suggestions can go in the acknowledgements? DONE

    I suggest common website links, like github.com, gitlab.com and cran be removed?
        Agreed.

    “Some other CRAN  packages…” can go later, after the main material of the paper?
        DONE -- before unit tests paragraph.

    Documentation and resources seem tangential to the main material and maybe removed?
        DONE -- this was when we were talking more about the GSoC project.

    Maybe add 5, 6 above in a separate r blog post or the main article’s supplementary article?
       ??? What supplementary article? If we want to reference blog posts, they should
       be available now. Perhaps set aside and focus on getting this published. Thoughts?

=========================================================================
Revew Responses to first review of R Journal paper 2023-15.

@@

Response 2:01

We appreciate the positive comments and agree with the suggested title change.

Response 2:02

Both referees have requested revisions to the introduction, and we hope that 
we have addressed the comments, essentially by clarifying the goals and how
they arose. In reconsidering the focus of the article, we have removed the 
"Documentation and Resources" section that was primarily reporting outputs 
of our Google Summer of Code project. We have also moved the section 
"Some other CRAN packages for nonlinear modeling" towards the end of the
paper.

@@

    We have added two small comments on recommended usage of bounds. One
    is to control for wild excursions in the optimization trajectory, as 
    we believe this is sensible when bounds constraints are available, 
    as in nlsr::nlxb. We also note that in cases where a simple and 
    obvious transformation is available, such as squaring a parameter to 
    ensure positivity, this is likely a better approach than imposing bounds.
    However, when there are more general bounds, transformations become more
    tricky to apply, and may de-scale a problem. We have considered the
    transfinite function used in CRAN package `adagio`, but consider that
    this needs more extensive investigation before recommending in general.

5. The section Programming language makes the argument that keeping to a 
single programming language can allow for easier maintenance and upgrades, 
and that the performance penalty for using code entirely in R has been 
much reduced in recent years. I agree that there is not much of a 
performance penalty when solving small problems, such as the data examples 
included in the article. However, when solving problems using a large amount 
of data or a large number of parameters, I have experienced that the run-time 
(as well as memory usage) can become an actual bottleneck. As an example, 
consider the Penalty function I problem (Moré, Garbow, and Hillstrom (1981)) 
included in the Illustrating code section.

Response 2:08  In that we are in general agreement with this comment, we have added
    some clarification to advise that it is worthwhile being aware of how long
    calculations are taking.

    The referee's commends and Illustrating Code led us to try the Pen1 problem
    (it is part of the funconstrain package https://github.com/jlmelville/funconstrain
    to which one of us recently added Hessian code). This showed that the R code
    of nlsr::nlfb() was generally slower by a small factor than the compiled 
    Fortran of minpack.lm::nls.lm. However, the all-R code optimx::Rcgmin was 
    much faster for larger problems -- it does not build a matrix!

@@

• Avoid emphasizing words by writing in all capitals. Examples: NOT (pg. 14, 15), 
WEIGHTED (pg. 13) and more.

Response 2:09  Fair enought. Done.

• The link https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211 (as well 
as the links to github) may be subject to change or become unavailable in the future.

Response 2:10  We agree with this opinion. 
    We have replaced the link with 'a number of blog posts found by searching "unit 
    testing in R"' as there are several possibilities, of which we felt that of 
    André Müller was helpful.

@@

## Penalty function I [More, Garbow, Hillstrom 1981]
alpha <- 1e-5
n <- 1000
### residuals
fres <- function(x) {
c(sqrt(alpha) * (x - 1), sum(xˆ2) - 0.25)
}
### jacobian
fjac <- function(x) {
jj <- rbind(diag(sqrt(alpha), nrow = length(x)), 2 * t(x))
attr(jj, "gradient") <- jj
jj
}
bench::mark(
"large problem" = nlsr::nlfb(start = 1:n, resfn = fres, jacfn = fjac),
iterations = 1
)

References

Moré, Jorge J., Burton S. Garbow, and Kenneth E. Hillstrom. 1981. “Testing Unconstrained Optimization Software.” ACM Transactions on Mathematical Software 7: 17–41.

-----------------------------------------------------------------
Pen1test.R code

# Penalty function I [More, Garbow, Hillstrom 1981]
# sink("Pen1Times.txt", split=TRUE)
alpha <- 1e-5
a <- 1e-5
sqrta <- sqrt(alpha)
### residuals
fres <- function(x) {
  c(sqrt(alpha) * (x - 1), sum(x^2) - 0.25)
}
### jacobian
fjac <- function(x) {
  jj <- rbind(diag(sqrt(alpha), nrow = length(x)), 2 * t(x))
  attr(jj, "gradient") <- jj
  jj
}

ffn = function(par) {
  n <- length(par)
  if (n < 1) {
    stop("Penalty Function I: n must be positive")
  }
  fsum <- 0
  fn1 <- 0
  for (i in 1:n) {
    fi <- sqrta * (par[i] - 1)
    fsum <- fsum + fi * fi
    fn1 <- fn1 + par[i] * par[i]
  }

  fn1 <- fn1 - 0.25
  fsum <- fsum + fn1 * fn1
  fsum
}

fgr = function(par) {
  n <- length(par)
  if (n < 1) {
    stop("Penalty Function I: n must be positive")
  }
  grad <- rep(0, n)
  fn1 <- 0

  for (i in 1:n) {
    fi <- sqrta * (par[i] - 1)
    grad[i] <- grad[i] + 2 * sqrta * fi
    fn1 <- fn1 + par[i] * par[i]
  }
  fn1 <- fn1 - 0.25
  grad <- grad + 4 * par * fn1
  grad
}
n<-10
# ftest
x0<-1:n
cat("ffn=",ffn(x0),"\n")
cat("fres:")
fr<-fres(x0)
print(fr)
cat("sum(fr^2)=", sum(fr^2), "\n")
cat("fgr:")
print(fgr(x0))
cat("Jacobian:\n")
JJ<-fjac(x0)
print(JJ)
cat("grad from Jacobian:")
grj <- 2*as.vector(t(JJ)%*%fr)
print(grj)

library(nlsr)
library(minpack.lm)
library(optimx)

nvals <- c(50, 100, 150, 200, 250, 300)
#  nvals <- c(2, 10, 25, 50)
nn<-length(nvals)
TT <- matrix(NA, nrow=nn, ncol=4)
MM <- TT
for (i in 1:nn){
n <- nvals[i]
TT[i, 1] <- n
MM[i, 1] <- n
x0<-as.numeric(1:n)
tnlsr10<-bench::mark(nlsr10 <- nlsr::nlfb(start = x0, resfn = fres, jacfn = fjac),  iterations = 10, time_unit='ms')
pshort(nlsr10)
TT[i,2] <- tnlsr10$median
MM[i,2] <- tnlsr10$min
tnlslm10<-bench::mark(nlslm10 <- nls.lm(par = x0, fn = fres, jac = fjac),  iterations = 10, time_unit='ms')
pnlslm(nlslm10)
TT[i,3] <- tnlslm10$median
MM[i,3] <- tnlslm10$min
tcg10<-bench::mark(cg10 <- optimr(par = x0, fn = ffn, gr = fgr, method="Rcgmin"),  iterations = 10, time_unit='ms')
proptimr(cg10)
TT[i,4] <- tcg10$median
MM[i,4] <- tcg10$min
}
colnames(TT)<-c("n","nlfb","nls.lm","Rcgmin")
colnames(MM)<-c("n","nlfb","nls.lm","Rcgmin")
print(TT)
print(MM)
sink()

### 2023-07-15 RefactoringNLS/revision202307/StateOfRToolsForNLS.Rmd
minor updates

Our Google Summer of Code project "Improvements to `nls()`" considered the features 
and limitations of the R function `nls()` with the aim 
of improving and rationalizing R tools for nonlinear regression.
The rich features of `nls()` are weakened by several deficiencies and
inconsistencies such as a lack of stabilization of the Gauss-Newton solver.
Further considerations are the usability and  maintainability of
the code base that provides the functionality `nls()` claims to offer.
Various packages, in particular our `nlsr`, provide alternative capabilities.
We consider the differences in goals, approaches, and features of
different tools for nonlinear least squares modeling in R. Discussion of 
these matters is relevant to improving R generally as well as its 
nonlinear estimation tools.

@@

`nls()` is the tool for estimating nonlinear statistical models in base R, the 
primary software distribution
from the Comprehensive R Archive Network (*https://cran.r-project.org*), hereafter
abbreviated CRAN.
Users do not have to request that `nls()` be loaded.
The function dates to the 1980s and the work related to @bateswatts and others 
in S (see *https://en.wikipedia.org/wiki/S_%28programming_language%29*).

@@

- handles formulas that include R functions, even ones that call calculations in
  other programming languages
- allows data to be weighted or subset
- can estimate bound-constrained parameters
- provides a mechanism for handling partially linear models
- permits parameters to be indexed over a set of related data
- produces measures of variability (i.e., standard error estimates) for the 
estimated parameters
- has related profiling capabilities for exploring the likelihood surface as 
parameters are changed
- links to many pre-coded (`selfStart`) models that do not require
  initial parameter values.

@@

Our Google Summer of Code project for 2021, "Improvements to `nls()`"
had the goal of providing a version of the code that would address some
of the deficiencies discovered over the long period that the `nls()`
has been part of R. However, it was also critical to preserve legacy
functionality, and we did not achieve a result that provided desired
improvements and preserved legacy results. On the other hand, working
with `nls()` suggested ways our own `nlsr` package could be improved
in features, while also revealing how it has a very different internal
structure and design.

The primary goal of this paper is to present some of these issues,
with a limited consideration of some other tools that exist for 
nonlinear least squares calculations. A hope is that we will encourage
others to engage, possibly with us, to find ways to bring improvements.
We note that the referees' comments have been most helpful to focus
the discussion here in ways that illustrate the best aspects of open
source software development. We believe it is desirable to consolidate
tools where reasonable so that users have an easier task in choosing
which to use, but such consolidation demands careful and wide-ranging
discussion of options. It is much simpler to create new packages than
to bring together and simplify existing ones.

@@

Besides the base-R `nls()` function, we consider some packages that 
are available on the CRAN repository.
Of these, we will pay particular attention to **nlsr** (@nlsr2023manual),
**minpack.lm** (@minpacklm12), and **gslnls** (@gslnls23) which are
general nonlinear least-squares solvers. We will 
not pursue tools in the Bioconductor (@Bioconductor) 
collection, nor those on repositories such as 
GitHub (https://github.com) and Gitlab (https://about.gitlab.com).

>>>>

Our Google Summer of Code project "Improvements to `nls()`" considered the features 
and limitations of the R function `nls()` with the aim 
of improving and rationalizing R tools for nonlinear regression.
The rich features of `nls()` are weakened by several deficiencies and
inconsistencies such as a lack of stabilization of the Gauss-Newton solver.
Further considerations are the usability and  maintainability of
the code base that provides the functionality `nls()` claims to offer.
Various packages, in particular `nlsr` (@nlsr2023manual), provide alternative capabilities.
We consider the differences in goals, approaches, and features of
different tools for nonlinear least squares modeling in R. Discussion of 
these matters is relevant to improving R generally as well as its 
nonlinear estimation tools.

@@

`nls()` is the tool for estimating nonlinear statistical models in base R, the 
primary software distribution
from the Comprehensive R Archive Network (CRAN).
Users do not have to request that `nls()` be loaded.
The function dates to the 1980s and the work related to @bateswatts and others 
in the S programming language.

@@

- handles formulas that include R functions, even ones that call calculations in
  other programming languages;
- allows data to be weighted or subset;
- can estimate bound-constrained parameters;
- provides a mechanism for handling partially linear models;
- permits parameters to be indexed over a set of related data;
- produces measures of variability (i.e., standard error estimates) for the 
estimated parameters;
- has related profiling capabilities for exploring the likelihood surface as 
parameters are changed;
- links to many pre-coded (`selfStart`) models that do not require
  initial parameter values.

@@

Our Google Summer of Code project for 2021, "Improvements to `nls()`"
had the goal of providing a version of the code that would address some
of the deficiencies discovered over the long period that `nls()`
has been part of R. However, it was also critical to preserve legacy
functionality, and we did not achieve a result that provided desired
improvements and preserved legacy results simultaneously. On the other hand, working
with `nls()` suggested ways package `nlsr` could be improved
in features, while also revealing how it has a very different internal
structure and design.

The primary goal of this paper is to present some of these issues,
with a limited consideration of some other R tools that exist for 
nonlinear least squares calculations. We hope to encourage
others to engage, possibly with us, to find ways to bring improvements and
to consolidate tools where reasonable. We want to ease the task of users
in choosing which tools to use. Unfortunately, it is simpler to create 
new packages than to bring together and simplify existing ones.

@@

Besides the base-R `nls()` function, we consider some packages that 
are available on the CRAN repository. Of these, we will pay particular 
attention to **nlsr** (@nlsr2023manual),
**minpack.lm** (@minpacklm12), and **gslnls** (@gslnls23) which are
general nonlinear least-squares solvers. We will 
not pursue tools in the Bioconductor (@Bioconductor) 
collection, nor those on repositories such as 
GitHub and Gitlab.

The Hobbs weed infestation problem (@cnm79 [page 120]) is a growth curve modeling 
task that seems straightforward, but is quite nasty. Its very succinct statement 
provides the "short reproducible example" much requested on R mailing lists. 
This is a real problem from a field researcher, though the units of measurement
and background information are not available. The data and graph follow.

Let us consider there are $n$ parameters and $m$ residuals. 
The loss function is 
$$ S(p) = r' r = \sum_{i=1}^m { r_i^2 }$$
The gradient of $S(p)$ is
$$ g = 2 * J' r$$
where the Jacobian $J$ is given by elements
$$ J_{i,j} = \partial r_i / \partial p_j $$
and the Hessian is defined by elements
$$ H_{i,j} = \partial ^2 S(p) / {\partial p_i \partial p_j} $$
If we expand the Hessian for nonlinear least squares problems, we find
$$  H_{i,j}/2 = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}$$
Let us use $D_{i,j}$ for the elements of the second term of this expression.
What is generally called **Newton's method** for function minimization tries
to set the gradient to zero (to find a stationary point of the function $S(p)$).
This leads to **Newton's equation**
$$ H \delta  = -g $$
Given a set of parameters $p$, we solve this equation for $\delta$, adjust $p$
to $p + \delta$ and iterate, hopefully to converge on a solution. 
Applying this to a sum of squares problem gives
$$ H \delta / 2 = (J'J + D) \delta = - J' r$$
In this expression, only the elements of $D$ have second partial derivatives. 
Gauss, attempting to model planetary orbits, had small residuals, and noted
that these multiplied the second partial derivatives of $r$, so he approximated
$$ H / 2 \approx J' J$$
by assuming $D \approx 0$. This results in the Gauss-Newton method where we
solve 
$$ J' J \delta = - J' r$$
though we can avoid some loss of accuracy by not forming the inner product matrix
$J' J$ and solving the linear least squares matrix problem 
$$ J \delta \approx -r $$
by one of several matrix decomposition methods. On the other hand, when $m >> n$,
the matrix $J' J$ uses much less storage than $J$.

In reality, there are many problems where $D$ should not be ignored, but the work
to compute it precisely is considerable. Many work-arounds have been 
proposed, of which the Levenberg-Marquardt stabilization 
(@Levenberg1944, @Marquardt1963) is
the most commonly used. For convenience, we will use "Marquardt", as we believe he
first incorporated the ideas into a practical computer program.

```{r ex02set, echo=TRUE}
# model formulas
frmu <- weed ~ b1 / (1 + b2 * exp(-b3 * tt))
frms <- weed ~ 100* c1 / (1 + 10 * c2 * exp(-0.1 *c3 * tt))
frmt <- weed ~ Asym / (1 + exp((xmid - tt) / scal))
#
# Starting parameter sets
stu1<-c(b1 =1, b2 = 1, b3 = 1)
sts1<-c(c1 = 1, c2 = 1, c3 = 1)
stt1<-c(Asym = 1, xmid = 1, scal = 1)
```

```{r ex02fn, echo=TRUE}
# Logistic3U
hobbs.res  <-  function(x){ # scaled Hobbs weeds problem -- residual
  # This variant uses looping
  if(length(x) != 3) stop("hobbs.res -- parameter vector n!=3")
  y  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
           38.558, 50.156, 62.948, 75.995, 91.972)
  tt  <-  1:12
  res  <-  x[1] / (1 + x[2] * exp(-x[3] * tt)) - y
}

hobbs.jac  <-  function(x) { # scaled Hobbs weeds problem -- Jacobian
  jj  <-  matrix(0.0, 12, 3)
  tt  <-  1:12
  yy  <-  exp(-x[3] * tt)
  zz  <-  1.0 / (1 + x[2] * yy)
  jj[tt,1]   <-   zz
  jj[tt,2]   <-   -x[1] * zz * zz * yy
  jj[tt,3]   <-   x[1] * zz * zz * yy * x[2] * tt
  attr(jj, "gradient") <- jj
  jj
}
```

Using a formula specification was a principal advantage made
with `nls()` when it became available in S sometime in the 1980s. It uses a
Gauss-Newton (i.e., unstabilized) iteration with a step reduction line
search (@Hartley1961). This works very efficiently as long as $J$ is not ill-conditioned.
Below we see the default `nls()` algorithm does poorly on the example problem. 
To save page space, we use 1-line result display functions from package `nlsr`, 
namely `pnls()` and `pshort()`.

```{r ex02nls}
unls1<-try(nls(formula = frmu, start = stu1, data = weeddf))
snls1<-try(nls(formula = frms, start = sts1, data = weeddf))
tnls1<-try(nls(formula = frmt, start = stt1, data = weeddf))
```

Here we see the infamous "singular gradient" termination message of `nls()`. 
Technically, it is the Jacobian that is singular, but R does not distinguish scalar 
and vector functions, which yield partial derivatives that are a gradient or Jacobian 
respectively.

```{r ex02nlsport}
unls1p<-try(nls(formula = frmu, start = stu1, data = weeddf, algorithm = "port"))
pnls(unls1p)
snls1p<-try(nls(formula = frms, start = sts1, data = weeddf, algorithm = "port"))
pnls(snls1p)
tnls1p<-try(nls(formula = frmt, start = stt1, data = weeddf, algorithm = "port"))
pnls(tnls1p)
```

```{r ex02nlsr}
unlx1 <- try(nlxb(formula = frmu, start = stu1, data = weeddf))
print(unlx1) 
snlx1 <- try(nlxb(formula = frms, start = sts1, data = weeddf))
pshort(snlx1) # a short-form output
tnlx1 <- try(nlxb(formula = frmt, start = stt1, data = weeddf))
pshort(tnlx1) # alternatively print(tnlx1)
```

```{r ex02minpack}
unlm1 <- try(nlsLM(formula = frmu, start = stu1, data = weeddf))
pnls(unlm1)  # Short form of output
snlm1 <- try(nlsLM(formula = frms, start = sts1, data = weeddf))
pnls(snlm1)
tnlm1 <- try(nlsLM(formula = frmt, start = stt1, data = weeddf))
pnls(tnlm1) # short form to give sum of squares, else use summary(tnlm1)
```

```{r ex02gslnls}
library(gslnls)
ugslnls1 <- try(gsl_nls(fn = frmu, data = weeddf,  start = stu1))
pnls(ugslnls1) # to get sum of squares
sgslnls1 <- try(gsl_nls(fn = frms, data = weeddf,  start = sts1))
pnls(sgslnls1) # Use summary() to get display
tgslnls1 <- try(gsl_nls(fn = frmt, data = weeddf,  start = stt1))
pnls(tgslnls1) 
## Other algorithms
## tgslnls1a<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="lmaccel"))
## pnls(tgslnls1a) 
## tgslnls1d<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="dogleg"))
## pnls(tgslnls1d) 
## tgslnls1dd<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="ddogleg"))
## pnls(tgslnls1dd) 
## tgslnls1s2<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="subspace2D"))
## pnls(tgslnls1s2) 
```

While default `gsl_nls()` gives much of the same results as `nlsLM()`, uncommenting the
calls to the alternative algorithms shows it can get the global minimum with any of
these choices.

`nlsr::nlxb()` uses `print()` to output standard errors and 
singular values of the Jacobian (for diagnostic purposes). By contrast, 
`minpack.lm::nlsLM()` and `nls()`
use `summary()`, which does not display the sum of squares, while `print()` 
gives the sum of squares, but not the standard error of the residuals.
Such behaviour is consistent with other modeling functions such as `lm()`.

The importance of the  singular values is to allow us to gauge how "nearly singular" 
the Jacobian is at the solution, and the ratio of the smallest to largest of the 
singular values is a simple but effective measure. The ratios are
4.6641e-05 for Logistic3U, 0.021022 for Logistic3S, and 0.001055 for
Logistic3T; so Logistic3S is the "least singular".

We see that there are differences in detail, but the more important result is that
two out of three singular values are essentially 0. Our Jacobian is singular, and no 
method of the Gauss-Newton type should be able to continue. Indeed, from the 
parameters reported at this saddle point, `nlsr::nlxb()` cannot proceed.

```{r ex05, echo=TRUE}  
stspecial <- c(Asym = 35.532,  xmid = 43376,  scal = -2935.4)
badstart <- nlxb(formula = frmt, start = stspecial, data = weeddf)
print(badstart)
```

We also 
note that changing the start to
```
stt2 <- c(Asym=1, xmid=1, scal=2)
```
allows both `nlsLM` and `gsl_nls` to find the minimum. `gsl_nls` also succeeds
when the particular form of the Marquardt stabilization is changed from the
default by the control setting `scale = "levenberg"`.

We consider the code inside package `nlsr` to be competently coded, 
with the advantage over `nls()` of the Levenberg-Marquardt stabilization 
in common with `minpack.lm` and `glsnls`. The success rate is high,
but `nlsr` has its premature terminations too. However, it provides 
diagnostic information in the form of the singular values of the Jacobian 
that allows us to try alternative starts, algorithms or programs in cases 
where there are indications of early termination.

R users are likely to think almost always that the functions $r_i$ are all of
the same form, differing in the details of the data they use. However, the
famous Rosenbrock banana-shaped valley (@Rosenbrock1960) has 
$$ r_1(p) = (1 - p_1)$$
$$ r_2(p) = 10 (p_2 - {p_1}^2)$$

Thus, there can be problems where the idea of a "model" or a "prediction"
makes no sense, though the nonlinear least squares problem is still
well-posed. In our code, we still refer to the functions as "residuals",
though this is purely a convenience.

We illustrate how to solve nonlinear least squares problems using
a function to define the residual. Note that `gsl_nls()` requires
a vector `y` that is the expected value of what we have called the
residual, but in this case is actually the model. `gsl_nls` uses 
a numerical approximation for the Jacobian if the argument `jac` 
is missing. Note functions `nlsr::pnls()` and `nslr::pnlslm()` for a 
1-line display of the results.

```{r exhobbsfn, eval=TRUE}
hobnlfb <- nlfb(start = stu1, resfn = hobbs.res, jacfn = hobbs.jac)
pshort(hobnlfb) # use print(hobnlfb) for more detail
hobnlm <- nls.lm(par = stu1, fn = hobbs.res, jac = hobbs.jac)
pnlslm(hobnlm)  
hobgsln <- gsl_nls(start = stu1, fn = hobbs.res, y = rep(0,12))
pnls(hobgsln)
hobgsl <- gsl_nls(start = stu1, fn = hobbs.res, y = rep(0,12), jac = hobbs.jac)
pnls(hobgsl) # using analytic Jacobian
```

The output object of `nlxb()` is smaller than the class `nls` object returned
by `nls()`, `nlsLM()`, and `gsl_nls()`. Package `nlsr` emphasizes the solution 
of the nonlinear least squares problem rather than the estimation of a nonlinear 
model that fits or explains the data. The object of class `nls` allows for 
a number of specialized modeling and diagnostic extensions. For compatibility, 
the `nlsr` package has the function `wrapnlsr()`, for which `nlsr()` is an alias,
that is intended for problems specified as formulas.
This uses `nlxb()` to find good parameters, then calls `nls()` to return the class 
`nls` object. Unless particular modeling features are needed, the use of 
`wrapnlsr()` is unnecessary and wasteful of resources.

Additional termination tests can be used. `nlsr` has a **small sum of 
squares** test (`smallsstest`) that compares the latest evaluated sum of squared
(weighted) residuals to `e4` times the initial sum of squares, where
`e4 <- (100 * .Machine$double.eps) ^ 4` is approximately 2.43e-55.

Termination after what may be considered excessive computation is also important.
`nls()` stops after `maxiter` "iterations". The meaning of "iteration" may require an 
examination of the code. `nlsr` terminates when the number of residual or 
Jacobian evaluations (`feval` and `jeval`) exceed set limits, and we prefer larger 
limits (`jemax = 5000` is the default) than the `maxiter = 50` of `nls()` to avoid 
stopping early. This may cause overly long run-times for `nlsr`.

The complexity of this object is a challenge to users. Let us use 
`result` as the name of the returned object from `nls()`, `minpack.lm::nlsLM()` or,
`gslnls::gsl_nls()`. For example, the `data` return element is an R symbol. 
If the user's environment exists, which it may not, we could see this data 
via the syntax:
```
eval(parse(text = result$data))
```
However, since the evaluating environment might not exist, a better approach 
is to make the call to `nls()` 
with `model = TRUE`, which returns an element 
`model` which contains the data, and we can list its contents using
```
ls(result$model)
```
and if there is an element called `xdata`, then it can be accessed as
`result$model$xdata`.

Tools that produce a class `nls` output object create a rich set of functions
and structures that are then used in a variety of modeling tasks, including the
least squares solution. By contrast,
`nlsr` computes quantities as they are requested or needed, with additional
features in separate functions. For example, the singular values of the Jacobian 
are actually computed in the `print` and `summary` methods for the result. 
These two approaches
lead to different consequences for performance and how features are
provided. `nlsr` has antecedents in the methods of @cnm79, where storage for
data and programs was at a ridiculous premium in the small computers of the era.
Thus the code in `nlsr` is likely of value for workers to copy and modify
for customized tools, with code in R to support such uses.

Gauss-Newton/Marquardt methods all need a Jacobian matrix at each iteration. 
By default, `nlsr::nlxb()` will try to evaluate this using analytic 
expressions using symbolic and automatic differentiation tools. 
When using a formula specification of the model, `nls()`, `minpack.lm::nlsLM()`
and `gslnls::gsl_nls()` use a finite difference approximation to compute the
Jacobian, though `gsl_nls()` does have an option to attempt symbolic expressions.
Package `nlsr` provides, via appropriate calling syntax, four numeric approximation 
options for the Jacobian, with a further control `ndstep` for the size of the
step used in the approximation. These features are mainly to allow for investigation
of approximate derivatives. Regular users will rarely need them.

- Karl Schilling (private communication) provided an example where a model specified
  with the formula `y ~ a * (x ˆ b)` causes `nlsr::nlxb` to fail because the partial
  derivative w.r.t. `b` is `a * (x^b * log(x))`. If there is data for which `x = 0`,
  the derivative is undefined, but the model can be computed. In such cases, 
  we observed that `nls()` and `minpack.lm::nlsLM` found a solution through the 
  lucky accident that the approximate derivative can be evaluated.

Analytic Jacobian code can be provided to all the solvers discussed.
Most `selfStart` models that automatically provide starting parameters
also include such code. There is documentation in R
of `selfStart` models, but their construction is non-trivial. A number 
of such models are included with base R in `./src/library/stats/R/zzModels.R`,
with package `nlraa` (@MiguezNLRAA2021) providing a richer set. 
There are also some in the now-archived package `NRAIA`.
The Jacobian is provided via the `gradient` attribute of the "one-sided" formula
that defines each model, and these Jacobians are often the analytic forms.

In considering `selfStart` models, we noted that the base-R function `SSlogis`
is intended to solve problem **Logistic3T** above.  When this function is used
via `getInitial()` to  find
starting values, it actually calls `nls()` with the 'plinear' algorithm 
and finds a (full) solution. It then passes the solution coefficients to the default 
algorithm which converges in one iteration and builds the standard output. 
We found the implicit double-call a challenge to unravel. 
A more conventional approach to starting parameters, following the lead of @Ratkowsky1983, 
is in the function `SSlogisJN`, now part of the package `nlsr`.

Users may also want to profit from the Jacobian code of `selfStart` models 
but supply explicit starting values other than those suggested by `getInitial()`.
This does not appear to be possible with `nls()`. `nlsr::nlxb()` always requires 
starting parameters. We could get these by separately calling `getInitial()` to 
find them from the `selfStart` model, but could use other values. The `selfStart` 
model is also used in the formula for the `nlxb()` call.

We note that the analytic expressions for the Jacobian (`gradient`) in the `SSLogis` 
function and others save quantities by default in "hidden" variables, i.e., with 
names preceded by ".".  This behaviour is to avoid name 
conflicts with user quantities. Hidden variables are not displayed by the `ls()` 
command, so users need to remember to use `ls(all.names = TRUE)` to see which quantities
are defined, for example, in developing their own selfStart model via copy and edit.

- `nls()` can only impose bounds if the `algorithm = "port"` argument is 
used in the call. Unfortunately, the documentation warns us:

Examples of the use of bounds constraints for nonlinear regression modeling are
quite scarce. Traditionally, workers transform variables to ensure bounds are
satisfied, such as squaring a parameter, or using a parameter which is the $\log()$
of the actual parameter, so that its exponential is positive. These transformations
can create scaling issues, but seem to generally work.

Bounds can also be helpful to avoid unanticipated excursions of the minimization
trajectory into inadmissible parameter regions. Such bounds should not be too
tight, as this may hinder progress to an optimum by causing the methods to 
bump along the constraint in many cycles. Our aim is to avoid rubbish, not
specify where the solution will be, and we hope the solution is complicated
by having a parameter on a bound.

```{r ex10, echo=TRUE} 
# Start MUST be feasible i.e. on or within bounds
anlshob1b <- nls(frms, start = sts1, data = weeddf, lower = c(0,0,0),
             upper = c(2,6,3), algorithm = 'port')
pnls(anlshob1b) #  check the answer (short form)
# nlsLM seems not to work with bounds in this example
anlsLM1b <- nlsLM(frms, start = sts1, data = weeddf, lower = c(0,0,0), upper = c(2,6,3))
pnls(anlsLM1b)
# also no warning if starting out of bounds, but gets a good answer!!
st4<-c(c1 = 4, c2 = 4, c3 = 4)
anlsLMob <- nlsLM(frms, start = st4, data = weeddf, lower = c(0,0,0), upper = c(2,6,3))
pnls(anlsLMob)
# Try nlsr::nlxb()
anlx1b <- nlxb(frms, start = sts1, data = weeddf, lower = c(0,0,0), upper = c(2,6,3))
pshort(anlx1b)
```

```{r ex10m, echo=TRUE}
# Hobbsmaskx.R -- masks with formula specification of the problem
require(nlsr); require(minpack.lm); traceval <- FALSE
stu <- c(b1 = 200, b2 = 50, b3 = 0.3) # a default starting vector (named!)
sts <- c(c1 = 2, c2 = 5, c3 = 3) # a default scaled starting vector (named!)
# fix first parameter
anxbmsk1 <- try(nlxb(frmu, start = stu, data = weeddf, lower = c(200,0,0), 
			upper = c(200, 60, 3), trace=traceval))
print(anxbmsk1)
anlM1 <- try(nlsLM(frmu, start = stu, data = weeddf, lower = c(200,0,0), 
			upper = c(200, 60, 3), trace=traceval))
pnls(anlM1)
anlsmsk1 <- try(nls(frmu, start = stu, data = weeddf, lower = c(200,0,0), 
		upper = c(200, 60, 3),  algorithm="port", trace = traceval))
pnls(anlsmsk1)
# Hobbs scaled problem with bounds, formula specification
anlxmsks1 <- nlxb(frms, start = sts, data = weeddf, lower = c(2,0,0),
                  upper = c(2,6,30))
print(anlxmsks1)
anlshmsk1 <- nls(frms, start = sts, trace = traceval, data = weeddf, lower = c(2,0,0),
             upper = c(2,6,30), algorithm = 'port')
pnls(anlshmsk1)
anlsLMmsks1 <- nlsLM(frms, start = sts, data = weeddf, lower = c(2,0,0),
                 upper = c(2,6,30))
pnls(anlsLMmsks1)

# Test with all parameters masked
anlxmskall<- try(nlxb(frms, start = sts, data = weeddf, lower = sts, upper = sts))
print(anlxmskall)
```

All four major tools illustrated solve some variant of the Gauss-Newton 
equations. `nls()` uses a modification of an approach suggested by @Hartley1961,
while `nlsr`, `gslnls`, and `minpack.lm` use flavours of @Marquardt1963. `gslnls`
offers an accelerated Marquardt method and three alternative methods which we
have not investigated.
Control settings for `nlxb()` or `nlfb()` allow exploration of Hartley and Marquardt 
algorithm variants. In general, the Levenberg-Marquardt stabilization 
is important in obtaining 
solutions in methods of the Gauss-Newton family, as `nls()` terminates 
too frequently and unnecessarily with `singular gradient` errors. Successful
incorporation of Marquardt stabilization into `nls()` would be a major improvement.

## Programming language

An important choice made in developing `nlsr` was to code entirely within the R 
programming language. `nls()` uses a mix of R, C, and Fortran, as does `minpack.lm`. 
`gslnls` is an R wrapper to various C-language routines in the GNU Scientific 
Library (@GSL-manual).
Generally, keeping to a single programming language can allow for 
easier maintenance and upgrades. It also avoids some work when there are changes 
or upgrades to libraries for the non-R languages. 
R is usually considered slower 
than most computing environments because it keeps track of objects and 
because it is usually interpreted. In recent years, the 
performance penalty for using code entirely in R has been much reduced
with the just-in-time compiler and other improvements. All-R computation 
may now offer acceptable performance. In `nlsr`, the use of R may be a
smaller performance cost than the aggressive approach to a solution,
such as large computational limits for residuals and Jacobians.

There are problems where use of R will be slower. For example, a
"large" case of the Penalty 1 test problem of @More1981TUO will take
noticeable time to solve. Problems in large numbers of parameters
create large computational matrices for the Jacobian or its inner product,
which Gauss-Newton methods create and process. 
This is confirmed by comparing `nlsr::nlfb()` and `minpack.lm::nls.lm()` on
the Penalty 1 problem of various sizes, where the latter, which is 
compiled Fortran internally, was moderately faster. However, the method
"Rcgmin" from package `optimx` (@p-optimx) proved to be much 
faster than either, despite being entirely written in R, and is a much more
reliable conjugate gradient minimizer than method "CG" of `optim()`. One
of us created both, but now deprecates the latter. 
Timings of CG methods are more sensitive to small differences in
setup and arithmetic than Gauss-Newton methods, and the pattern of
such timings varies much less regularly with problem size for CG methods.
